---
title: "Figure 1b: Confounder analysis in drug-host-microbiome associations"
output:
  html_document:
    df_print: paged
---

Violin plot representing confounder analysis of features associated with T2D group compared to control.

Vertical axis shows Cliff’s delta, with density plots showing all features significant before drug deconfounding (two-sided MWU FDR < 0.1); number of features in each violin are listed. 

All “naïve associations” (yellow violin) are separated into confounded or ambiguously/confidently deconfounded (blue, purple and red violins). About 45% of 2845 raw associations are reducible to medication effects (blue violin). Green violins show breakdown of significant drug confounders by drug category. 

Required file in the _input_data_ folder:

- Supplementary_Table_5_2019-09-13434.xlsx


Load necessary libraries.
```{r, warning=FALSE}
library (ggplot2)
library (readxl)
```

Define accessory function. 
```{r}
f <- function (y) c (label = length (y), y = median (y))
```

Read input data from Supplementary Table 5. 
```{r}
fileFolder = './input_data/'
data <- read_excel(paste0(fileFolder,
                             "Supplementary_Table_5_2019-09-13434.xlsx"), 
                      sheet = "Confounder analysis")
```

Prepare data for plotting. 
```{r}
data$StatusD <- paste (data$Status, as.character (sign (data$`Effect size`)))

# Get unique feature spaces
spaces <- unique(data$`Feature space`)

data <- subset (data,`Feature space` %in% spaces)

# Get unique contrasts (patient groups)
contrasts <- unique(data$Contrast)

#Get unique confounder status
stati <- unique(data$Status)

data <- subset (data, Status %in% stati)
```

Plot violins for T2D group. 

*Change group index or loop over all contrasts to plot for a different patient group.*

```{r fig.width=20, fig.height=6, warning=FALSE}
for (i in c(4)) { #for (i in 1:7)

	subdata <- subset (data, Contrast == contrasts [i])

	p <- ggplot (subdata, aes (x = as.factor (Status), y = `Effect size`)) + #theme_classic () +
	  scale_x_discrete(limits = rev(levels(as.factor (data$Status)))) + 
	  geom_violin (aes (fill = Status), scale = "count", draw_quantiles = c (0.25, 0.5, 0.75), size = 0.5, alpha = 0.3, width = 30, position = position_dodge ()) + 
	  geom_hline (yintercept = 0) + 
	  xlab ("") + 
	  ylab ("Effect size (Cliff's delta)") +
	  ggtitle(contrasts[i]) + 
	  stat_summary (aes (col = subdata$Status, y = - 1), fun.data = f, geom = "text", vjust = 0, size = 4, angle = 45) + 
	  theme (axis.text.x = element_text (angle = -90, hjust = 0, size = 12), 
	         axis.text.y = element_text (size = 16), 
	         axis.title.y = element_text (size = 16), legend.position = "none") +
	  scale_fill_manual (values = c ("#00FF00", "#00FF00", "#00FF00", "#00FF00", "#00FF00", "#00FF00", "#0000FF", "#FF00FF", "#FF0000", "#DDCC00")) + 
	  scale_color_manual (values = c ("#009900", "#009900", "#009900", "#009900", "#009900", "#009900", "#0000FF", "#990099", "#990000", "#998800")) 

	#Uncomment to print to file
	#pdf (file = paste0 ("fig1b_flame_", i, ".pdf"), width = 20, height = 6)
	print (p)
	#dev.off ()

}

```


```{r}
sessionInfo()
```


