---
title: "Figure 3abc: Associations between long-term antibiotic exposure and microbiome"
output:
  html_document:
    df_print: paged
---

This script plots Figure 3abc: Additive drug effects on the microbiome features.

Scatterplots show microbiome features (a. Gene richness; b. Total abundance of antibiotic resistance genes; c. The first principal component of gut species composition) significantly associated with the number of antibiotics courses in the last 5 years in control and T2D subjects separately (with lines and gray area 95% CI for linear regression). 

Boxplots (box showing median and quartiles, whiskers 1.5 interquartile range) show the comparisons in antibiotics-naïve and antibiotics-exposed control and T2D subjects, respectively, with pairwise significances (two-sided MWU tests, FDR-adjusted).

Required file in the _input_data_ folder:

- Supplementary_Tables_1_to_4_2019-09-13434.xlsx
- hub.microbiome.summary.down.10000000.r
- hub.motus.SpeciesCluster.down.10000000.r


Load necessary libraries. 
```{r, warning=FALSE, message=FALSE}
library (factoextra)
library (vegan)
library (reshape2)
library (gtools)
library (ggplot2)
library (orddom)
library (cowplot)
library (readxl)
```

Read antibiotic course data from Supplementary Table 2d. 
```{r}
fileFolder = './input_data/'
md <- read_excel(paste0(fileFolder,
                             "Supplementary_Tables_1_to_4_2019-09-13434.xlsx"), 
                      sheet = "ST 2d")
```

Leave only patients with information on antibiotic courses (column ANTIBIOTICS_TOTAL)
```{r}
md <- subset (md, ! is.na (ANTIBIOTICS_TOTAL))
md <- as.data.frame(md)
row.names (md) <- md$SampleID
```

Get microbiome feature data from files.
```{r}
ssd <- read.table (file = paste0(fileFolder,
                                 "hub.microbiome.summary.down.10000000.r"),
                   header = T, check.names = F, sep = "\t")
ssdata <- dcast (ssd, value.var = "FeatureValue", formula = SampleID ~ Feature)
row.names (ssdata) <- ssdata$SampleID
```

Keep only samples with both antibiotic and microbiome data.
```{r}
md <- subset (md, SampleID %in% intersect (row.names (md), row.names (ssdata)))
ssdata <- subset (ssdata, SampleID %in% intersect (row.names (md), row.names (ssdata)))

md <- merge (ssdata, md, by = "SampleID")
row.names (md) <- md$SampleID

```

Restrict analysis to diabetics without AB exposure and to controls

```{r}
comp1 <- subset (md, (ANTIBIOTICS_TOTAL == 0 & PatientGroup == "3") | PatientGroup == "8")$SampleID

```

Load metagenome data from file. 

```{r}
data <- read.table (file = paste0(fileFolder, 
                                  "hub.motus.SpeciesCluster.down.10000000.r"),
                    header = T, sep = "\t")
data <- subset (data, SampleID %in% row.names (md))

# subset samples from the first comparison
data1 <- subset (data, SampleID %in% comp1)

#reshape into matrices
ddata <- dcast (data, formula = SampleID ~ Feature, value.var = "FeatureValue")
ddata1 <- dcast (data, formula = SampleID ~ Feature, value.var = "FeatureValue")
row.names (ddata) <- ddata$SampleID
row.names (ddata1) <- ddata1$SampleID
ddata <- ddata [, -1]
ddata1 <- ddata1 [, -1]
ddata <- na.omit (ddata)
ddata1 <- na.omit (ddata1)

md <- subset (md, SampleID %in% row.names (ddata))

```

Perform PCA analysis

```{r}
cds <- prcomp (ddata1, retx = T)

# project all samples into that space

ccds <- as.matrix (ddata) %*% cds$rotation

ccds <- merge (ccds, md, by = 0)
```

Testing AB associations

```{r, warning=FALSE}
# corr between AB courses and PC in controls
ccdssub1 <- subset (ccds, PatientGroup == "8")
cor.test (ccdssub1$PC1, ccdssub1$ANTIBIOTICS_TOTAL, method = "spearman")
cor.test (ccdssub1$PC2, ccdssub1$ANTIBIOTICS_TOTAL, method = "spearman")

# corr between AB courses and PC in T2D patients
ccdssub2 <- subset (ccds, PatientGroup == "3")
cor.test (ccdssub2$PC1, ccdssub2$ANTIBIOTICS_TOTAL, method = "spearman")
cor.test (ccdssub2$PC2, ccdssub2$ANTIBIOTICS_TOTAL, method = "spearman")
```

Perform Wilcoxon test for principal components between patient and control groups on ABX and not. 
```{r}
s8na <- subset (ccds, ANTIBIOTICS_TOTAL == 0 & PatientGroup == "8")
s8a <- subset (ccds, ANTIBIOTICS_TOTAL > 0 & PatientGroup == "8")
s3na <- subset (ccds, ANTIBIOTICS_TOTAL == 0 & PatientGroup == "3")
s3a <- subset (ccds, ANTIBIOTICS_TOTAL > 0 & PatientGroup == "3")

wilcox.test (s8na$PC2, s8a$PC2)
wilcox.test (s3na$PC2, s3a$PC2)
wilcox.test (s8na$PC2, s3na$PC2)
wilcox.test (s8a$PC2, s3a$PC2)
wilcox.test (s8na$PC2, s3a$PC2)
wilcox.test (s8a$PC2, s3na$PC2)
```

Perform Wilcoxon test for gene richness between patient and control groups on ABX and not. 
```{r}
wilcox.test (s8na$GeneCount10M, s8a$GeneCount10M)
wilcox.test (s3na$GeneCount10M, s3a$GeneCount10M)
wilcox.test (s8na$GeneCount10M, s3na$GeneCount10M)
wilcox.test (s8a$GeneCount10M, s3a$GeneCount10M)
wilcox.test (s8na$GeneCount10M, s3a$GeneCount10M)
wilcox.test (s8a$GeneCount10M, s3na$GeneCount10M)
```

Perform Wilcoxon test for CARD antibiotic resistance genes between patient and control groups on ABX and not. 
```{r}
wilcox.test (s8na$CARD10M, s8a$CARD10M)
wilcox.test (s3na$CARD10M, s3a$CARD10M)
wilcox.test (s8na$CARD10M, s3na$CARD10M)
wilcox.test (s8a$CARD10M, s3a$CARD10M)
wilcox.test (s8na$CARD10M, s3a$CARD10M)
wilcox.test (s8a$CARD10M, s3na$CARD10M)
```

```{r}
ccdssub2 <- subset (ccds, ANTIBIOTICS_TOTAL == 0 & (PatientGroup == "8" | PatientGroup == "3"))
wilcox.test (ccdssub2$PC1 ~ ccdssub2$PatientGroup)
wilcox.test (ccdssub2$PC2 ~ ccdssub2$PatientGroup)
orddom (subset (ccdssub2, PatientGroup == "8")$PC2, subset (ccdssub2, PatientGroup == "3")$PC2) [13]

ccdssub3 <- subset (ccds, PatientGroup == "8" | PatientGroup == "3")
ccdssub3$Status <- paste (as.character (10 - as.numeric (as.character(ccdssub3$PatientGroup))), as.numeric (ccdssub3$ANTIBIOTICS_TOTAL > 0))
wilcox.test (ccdssub3$PC2 ~ ccdssub3$PatientGroup)
orddom (subset (ccdssub3, PatientGroup == "8")$PC2, subset (ccdssub3, PatientGroup == "3")$PC2) [13]
```

Plot AB courses vs PC, diversity, ARG abundance
```{r}
p_abx_pc1 <- ggplot (ccdssub3, aes (x = ANTIBIOTICS_TOTAL, y = PC1)) + 
  theme_classic () + theme (legend.position = "none") + 
  geom_point (size = 0.5, alpha = 0.5, aes (colour = as.factor (PatientGroup))) + 
  geom_smooth (method = "lm", aes (colour = as.factor (PatientGroup))) + 
  coord_cartesian (ylim = c (- 400, 3200)) 

p_t2d_pc1 <- ggplot (ccdssub3, aes (x = as.factor (Status), y = PC1)) + 
  theme_classic () + theme (legend.position = "none") + 
  geom_boxplot (aes (fill = as.factor (PatientGroup))) + 
  coord_cartesian (ylim = c (- 400, 3200)) + 
  scale_x_discrete(labels= c('CTR noAB', 'CTR AB', 'T2D noAB', 'T2D AB'))

p_abx_pc2 <- ggplot (ccdssub3, aes (x = ANTIBIOTICS_TOTAL, y = PC2)) + 
  theme_classic () + theme (legend.position = "none") + 
  geom_point (size = 0.5, alpha = 0.5, aes (colour = as.factor (PatientGroup))) + 
  geom_smooth (method = "lm", aes (colour = as.factor (PatientGroup))) + 
  coord_cartesian (ylim = c (- 400, 3200))

p_t2d_pc2 <- ggplot (ccdssub3, aes (x = as.factor (Status), y = PC2)) + 
  theme_classic () + theme (legend.position = "none") + 
  geom_boxplot (aes (fill = as.factor (PatientGroup))) + 
  coord_cartesian (ylim = c (- 400, 3200)) + 
  scale_x_discrete(labels= c('CTR noAB', 'CTR AB', 'T2D noAB', 'T2D AB'))

p_abx_gc <- ggplot (ccdssub3, aes (x = ANTIBIOTICS_TOTAL, y = GeneCount10M)) + 
  theme_classic () + theme (legend.position = "none") + 
  geom_point (size = 0.5, alpha = 0.5, aes (colour = as.factor (PatientGroup))) + 
  geom_smooth (method = "lm", aes (colour = as.factor (PatientGroup))) + 
  coord_cartesian (ylim = c (100000, 1000000))

p_t2d_gc <- ggplot (ccdssub3, aes (x = as.factor (Status), y = GeneCount10M)) + 
  theme_classic () + theme (legend.position = "none") + 
  geom_boxplot (aes (fill = as.factor (PatientGroup))) + 
  coord_cartesian (ylim = c (100000, 1000000)) + 
  scale_x_discrete(labels= c('CTR noAB', 'CTR AB', 'T2D noAB', 'T2D AB'))

p_abx_abx <- ggplot (ccdssub3, aes (x = ANTIBIOTICS_TOTAL, y = CARD10M)) + 
  theme_classic () + theme (legend.position = "none") + 
  geom_point (size = 0.5, alpha = 0.5, aes (colour = as.factor (PatientGroup))) + 
  geom_smooth (method = "lm", aes (colour = as.factor (PatientGroup))) + 
  coord_cartesian (ylim = c (85000, 165000)) 

p_t2d_abx <- ggplot (ccdssub3, aes (x = as.factor (Status), y = CARD10M)) + 
  theme_classic () + theme (legend.position = "none") + 
  geom_boxplot (aes (fill = as.factor (PatientGroup))) + 
  #geom_jitter(color="black", size=0.4, alpha=0.9) +
  coord_cartesian (ylim = c (85000, 165000)) + 
  scale_x_discrete(labels= c('CTR noAB', 'CTR AB', 'T2D noAB', 'T2D AB'))
```

Make the plot

```{r, fig.width=6, fig.height=10, warning=FALSE}
pg <- plot_grid (p_abx_gc, p_t2d_gc, p_abx_abx, p_t2d_abx, p_abx_pc2, p_t2d_pc2,
                 ncol = 2, nrow = 3, 
                 labels = c ("", "", "", "", "", ""), align = "v")

print (pg)
# UNCOMMENT TO SAVE FIGURE TO FILE
#pdf (file = "fig3abc_abx_courses_associations.pdf", width = 6, height = 10)
#print (pg)
#dev.off ()
```








