---
title: "Figure 2b: Drug combinations graph"
output:
  html_document:
    df_print: paged
---

This script prints Figure 2b: The thirty most common drug combinations represented as a graph.

Node size is proportional to the number of combinations per drug; combinations with more than 3 drugs are omitted for better visualization.

Drugs are represented in nodes coloured by drug indication. Drug combinations are shown in the network links. Combination pairs are represented by solid lines; triad combinations are represented by distinct dotted/dashed lines. Link width is proportional to the number of users per combination; colour corresponds to the number of potential drug effects.
Combinations not tested are shown in gray.

Required file in the _input_data_ folder:

- Supplementary_Tables_1_to_4_2019-09-13434.xlsx
- Supplementary_Table_8_2019-09-13434.xlsx
- drug.indication.tsv
- data_drugs_dict.txt


Load required libraries.
```{r, message=FALSE, warning=FALSE}
set.seed(0)
library(reshape2)
library(stringr)
library(dplyr)
library(arules)
library(arulesViz)
library(igraph)
library(ggplot2)
library(RColorBrewer)
library(checkmate)
library(readxl)
```


Load data required for plotting from files.
```{r}
folderName <- './input_data/'
d.out <- "./"
drug.indication <- paste(folderName, "drug.indication.tsv", sep="")
dic.anno <- paste(folderName, "data_drugs_dict.txt", sep="")

dir.create(d.out, showWarnings = F)

infilename <- 'Supplementary_Tables_1_to_4_2019-09-13434.xlsx'
# load patient drug uptake information
sampleData_drug <- as.matrix(read_excel(paste(folderName,infilename,sep=''),
                                 sheet = "ST 2b"))#, 
# load patient metadata information
sampleData_meta <- as.matrix(read_excel(paste(folderName,infilename,sep=''),
                                        sheet = "ST 1b"))#, 
# load drug combination intake information
sampleData_comb <- as.matrix(read_excel(paste(folderName,infilename,sep=''),
                                        sheet = "ST 3b"))#, 

```

Define accessory functions.
```{r}
add.alpha <- function(col, alpha=1){
  if(missing(col))
    stop("Please provide a vector of colours.")
  apply(sapply(col, col2rgb)/255, 2,
        function(x)
          rgb(x[1], x[2], x[3], alpha=alpha))
}
```

Calculate the number of associations per drug from Supplementary Table 8
```{r}
infilename <- 'Supplementary_Table_8_2019-09-13434.xlsx'

combData <- as.matrix(read_excel(paste(folderName,infilename,sep=''),
                                 sheet = "Data"))#, 

combData_unique <- unique((combData[,"Effector"]))

effects_all <- rep(0,length(combData_unique))
effects_same <- rep(0,length(combData_unique))
effects_opp <- rep(0,length(combData_unique))

first_array <- rep('NA',length(combData_unique))
second_array <- rep('NA',length(combData_unique))
third_array <- rep('NA',length(combData_unique))

for (i in 1:length(combData_unique)){
  test <- combData[combData[,'Effector']==combData_unique[i],'Congruence']
  effects_all[i] = length(test)
  effects_opp[i] = sum(test=='Opposite')
  effects_same[i] = sum(test=='Same')
  curdrugs = str_split(combData_unique[i],'intake', simplify = TRUE)
  first_drug = curdrugs[1]
  first_drug = str_replace(first_drug, 'Combination: ', '')
  first_drug = str_trim(first_drug)
  second_drug = curdrugs[2]
  second_drug = str_replace(second_drug, ', ', '')
  second_drug = str_trim(second_drug)

  first_array[i] = first_drug
  second_array[i] = second_drug

  if (length(curdrugs)>3){
    third_drug = curdrugs[3]
    third_drug = str_replace(third_drug, ', ', '')
    third_drug = str_trim(third_drug)
    third_array[i] = third_drug
  }
}

synergy <- data.frame(first_array, second_array, third_array,
                              effects_all, effects_same, effects_opp)
colnames(synergy) <- c("Drug1",	"Drug2", "Drug3",
                              "AnySynergy",	"SeverityMarker",	"HealthImprovement")
#replace text NA with numeric NA
synergy[synergy=='NA']=NA
```



Prepare data for processing.
```{r}
# prepare drug table
rownames(sampleData_drug) = sampleData_drug[, "SampleID"]
drop <- c( "SampleID" ,"PatientGroup","DRUGTOTAL")
sampleData_drug = sampleData_drug[,!(colnames(sampleData_drug) %in% drop)]

# prepare drug comb table
rownames(sampleData_comb) <- sampleData_comb[, "SampleID"]
drop <- c( "SampleID" ,"PatientGroup")
sampleData_comb = sampleData_comb[,!(colnames(sampleData_comb) %in% drop)]

# prepare metadata
colnames(sampleData_meta)[2:6]<-c("PATGROUPFINAL_C", "GENDER", "BMI_C", "AGE", "CENTER_C")
rownames(sampleData_meta) <- sampleData_meta[, "SampleID"]

sample_extended_meta <- merge(sampleData_meta, sampleData_comb, by="row.names", all=TRUE)  
drop <- c( "Row.names")
sample_extended_meta = sample_extended_meta[,!(colnames(sample_extended_meta) %in% drop)]
```

Load data files for indication and drug abbreviations.
```{r}
# load drug abbreviation dictionary
dic.anno <- read.delim(dic.anno, stringsAsFactors = F)

# load rug indication info
drug.indication = read.delim(drug.indication, stringsAsFactors = F)

drugs <- sampleData_drug
metadata <- sample_extended_meta
```


Edit Dictionary.
```{r}
dic.anno <- dic.anno %>% 
  mutate(DisplayName = str_replace(DisplayName, " intake", ""))

```

Edit input data.
```{r}
## Filter out healthy patients
metadata <- metadata[ metadata$PATGROUPFINAL_C != "8",]
drugs <- drugs[match(metadata$SampleID, rownames(drugs)) ,]
identical(metadata$SampleID %>% as.character(), rownames(drugs))

## Remove collective drugs 
drugs <- drugs [, !colnames(drugs) %in% 
                  dic.anno[dic.anno$DisplayName %in%
                             c("Antibiotics total",
                               "Unknown antibiotic",
                               "Any antilipid treatment",
                               "Any antihypertensive treatment",
                               "Any antidiabetic treatment"),
                           "VariableID"]]
```


Sync data with dictionary.
```{r}
drugs <- drugs[,colnames(drugs) %in% dic.anno$VariableID]
dic.anno <- dic.anno[match(colnames(drugs), dic.anno$VariableID),]
identical(dic.anno$VariableID, colnames(drugs))

```

Format data
```{r}
## Rename columns to be the display variable
colnames(drugs) <- dic.anno$DisplayName

## Convert from factor to numeric
#for (i in 1:ncol(drugs)){
#  drugs[,i] <- as.numeric(drugs[,i])-1
#}
# Get frequent items ----
## Format data input
drugs.d <- drugs
drugs.d <- as.data.frame(drugs.d > 0)

## Mine frequent itemsets with the Eclat algorithm
freq.all=eclat(drugs.d,parameter=list(supp = 0.05))
# ## Get clean data
freq.list <- utils::capture.output(inspect(freq.all))
test = ''
for(i in 1:length(freq.list)){
  test<-paste(test, freq.list[i],sep=' ')
  }
freq.list <- strsplit(test, '\\{[^}]+ (*SKIP)(*FAIL)| \\s*', perl=TRUE)[1]
dd  <-  as.data.frame(t(matrix(unlist(freq.list), nrow=5, byrow=FALSE)))
dd <- dd[,-1]
names(dd) <- as.character(unlist(dd[1,]))
dd <- dd[-1,]
freq.list <- dd[-1,]
## End of Get clean data
# note: if running this code in console, instead of the code in Get clean data use
#freq.list <- as.data.frame(inspect(freq.all))
freq.list$items <- gsub("\\{", "", freq.list$items)
freq.list$items <- gsub("\\}", "", freq.list$items)
freq.list$items <- gsub("\\s+", " ", freq.list$items)
freq.list$items <- gsub(", ", ",", freq.list$items)
freq.list$support <- as.numeric(as.character(freq.list$support))
freq.list$transIdenticalToItemsets <- as.numeric(as.character(freq.list$transIdenticalToItemsets))
freq.list$count <- as.numeric(as.character(freq.list$count))
```

Write raw output.
```{r}
#UNCOMMENT TO PRINT FREQUENCIES TO FILE
#f.name <- paste(d.out, "frequent.list.5.tsv", sep = "")
#write.table(freq.list, f.name, row.names = F, sep = "\t", quote =F )
```


Annotate number of drugs in freq-items co-usage.
```{r}
freq.list <- freq.list %>%
  mutate (n = str_count(items, ",") + 1)
```

Filter what to plot in the network.
```{r}
to.net <- freq.list %>%
  # Keep combinations with up to 3 drugs
  filter(n <= 3, n > 1)

to.net$id = 0
synergy$id = 1:nrow(synergy)
```


Identify which freq items have synergy information by common id.
```{r}
for (x in 1:nrow(to.net)){
  set1 = to.net[x, "items"]
  set1 = str_split(set1, ",")[[1]]
  set1 = str_trim((set1))
  for (y in 1:nrow(synergy)){
    set2 = synergy[y,c("Drug1", "Drug2", "Drug3")]
    set2 = set2[!is.na(set2)]
    if(testSetEqual(set1, set2, ordered = F)){to.net$id[x]=synergy$id[y]}
    }
}
```

Filter network by count and add synergy information.
```{r}
top = 30

to.net <- synergy %>%
  select(-Drug1, -Drug2, -Drug3) %>%
  right_join(to.net) %>%
  top_n(n = top, wt = count)
```


```{r, warning=FALSE}
# Define cbind/fill function as in https://gist.github.com/abelsonlive/4112423
cbind.fill<-function(...){
    nm <- list(...)
    nm<-lapply(nm, as.matrix)
    n <- max(sapply(nm, nrow))
    do.call(cbind, lapply(nm, function (x)
    rbind(x, matrix(, n-nrow(x), ncol(x)))))
}
```

Format it to plot.
```{r}
## Get from and to edges
df.convoluted <- NULL
x <- 0
for (i in 1:nrow(to.net)){
its <- as.character(to.net[i,"items"])
its <-  str_split(its, ",")[[1]]
n <- length(its)
x <- x+1
if (n > 1) {
  its <- cbind(t(combn(its, 2)), code = x) %>%
    data.frame(stringsAsFactors = F)
  colnames(its) <- c("from", "to", "code")
  its <- as.data.frame(cbind.fill(its, to.net[i,]))
  df.convoluted <- rbind(df.convoluted,
                         its)
  }
}

## Transform info as character

df.convoluted <- df.convoluted %>%
  mutate(from = as.character(from),
         to =  as.character(to),
         code = as.character(code))

## Add edge with
x = 1
df.convoluted$edge.lty <- 1
for (i in unique(df.convoluted$code)){
  d.temp <- df.convoluted[df.convoluted$code == i,]
  if(nrow(d.temp) > 1){x <- x+1
  df.convoluted[df.convoluted$code == i, "edge.lty"] <- x}
}

## Vertex width
vert <- c(df.convoluted$from, df.convoluted$to) %>%
  unique() %>%
  sort() %>%
  data.frame(stringsAsFactors = F)
colnames(vert) = "vert"
vert <- vert %>%
  left_join(freq.list, by = c("vert" = "items")) %>%
  left_join(drug.indication,
            by = c("vert" = "Drug"))
```


Build network.
```{r}
net  <- graph_from_data_frame(d = df.convoluted, vertices = vert, directed=F)
```


Make cosmetic additions to it.
```{r}
## Make gradient of colors
rbPal <- colorRampPalette(c("#43C6AC","#191654"))
df.convoluted$Col <- add.alpha(rbPal(10)[
  as.numeric(cut(as.numeric(as.character(df.convoluted$HealthImprovement)), breaks = 10))], 0.8)
# #Mod color of NAs
col.2 <- add.alpha("#545454", 0.8)
df.convoluted <- df.convoluted %>%
  mutate(Col = if_else(is.na(HealthImprovement), col.2, Col))
```
```{r}
## Colors of nodes
n.col = length(unique(vert$Indication))
pal <- brewer.pal(n = 8, name = "Dark2")[c(1:3,5:6)]
pal <-  colorRampPalette(pal)(n.col)

## Get degrees
deg <- igraph::degree(net, mode="all")
```
```{r}
## Config vertex
V(net)$size=sqrt(deg-0.20)*9
V(net)$label.family="Helvetica"
V(net)$frame.color = NA
V(net)$label.color = "black"
V(net)$color = pal[as.numeric(factor(vert$Indication))]
V(net)$label.cex <- 1.2
```

```{r}
## Config edge
E(net)$label.dist=-2
E(net)$label.cex = 0.9
E(net)$label.colour = rep("black", nrow(df.convoluted))

E(net)$width = as.numeric(as.character(df.convoluted$count))/40

E(net)$color <- df.convoluted$Col
E(net)$label.color = "darkgray"

E(net)$lty <- df.convoluted$edge.lty
l <- layout_with_kk(net)
l <- layout_in_circle(net)
```



#Plot
```{r, warning=FALSE}
par(mar=c(0,0,0,0))
plot(net, layout=l,
    edge.label.color = "darkgrey")#edge.label.family = "Arial")
```

```{r}
## UNCOMMENT TO PRINT TO FILE
#f.name <- paste(d.out, "fig2b_network.pdf", sep ="/")
#pdf(f.name)
#par(mar=c(0,0,0,0))
#plot(net, layout=l,
#    edge.label.color = "darkgrey",
#    edge.label.family = "Helvetica")
#dev.off()
```


Plot legends separately
```{r}
## Legend colors vertex
vert.unique <- vert %>%
  select(n, Indication) %>%
  unique()

#f.name <- paste(d.out, "network.legend.1.pdf", sep ="/")
#pdf(f.name)
barplot(rep(5, nrow(vert.unique)), col = pal, legend.text = vert.unique$Indication %>% sort())
#dev.off()
```


Legend colors edge
```{r}
values  <- seq(from = min(as.numeric(as.character(df.convoluted$HealthImprovement)), na.rm = T),
    to = max(as.numeric(as.character(df.convoluted$HealthImprovement)), na.rm = T))
rbPal <- colorRampPalette(c("#43C6AC","#191654"))
col.legend <- add.alpha(rbPal(10)[
  as.numeric(cut(values, breaks = 10))], 0.8)
col.legend <- data.frame(values, col.legend, stringsAsFactors = F)
col.legend$v <- rep(1, nrow(col.legend))

#f.name <- paste(d.out, "network.legend.2.pdf", sep ="/")
#pdf(f.name)

barplot(as.matrix(data.frame(b = col.legend$v)),
        col = col.legend$col.legend, ylim = c(0, 90), beside = F,
        border=NA, space=0, xlab = 'Number of potential drug effects')

#dev.off()


```

```{r}
sessionInfo()
```

